Clustering time series based on probability distributions across temporal granularities

### 2021-11-02 main.Rmd
Do not embellish the plots mormore before Thursday

@@

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature, yet it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection.
Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers who are far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that was hidden in a single variable display better than static projections. Hence we perform a linked tour with a t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to Supplementary article for further details). Figure \ref{fig:hod-ind-group} (a, b, c) shows the raw time plot, distribution across `hod`, `moy` and `wnwd` for the set of chosen $24$ customers. Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution.

```{r prototype-data-pick}
quantile_prob_graph <- c(0.25, 0.5, 0.75)

# data_pick_one <- c(8618759, 8291696, 10357256, 8290374) %>% as_tibble %>% set_names("customer_id")
# data_pick_two <- c(9044864, 8642053, 10534367, 9021526,11162275) %>% as_tibble %>% set_names("customer_id")
# data_pick_three <- c(8221762, 8273636, 10359424, 8232822)%>% as_tibble %>% set_names("customer_id")

data_pick_one <- c(8541744, 9355808, 8603880, 8619309, 10542667) %>% as_tibble %>% set_names("customer_id")
#data_pick_two <- c(8688242, 8643837, 8184707, 10534355, 8684420) %>% as_tibble%>% set_names("customer_id")
data_pick_three <- c(9792072, 8589936, 8454235, 10692366, 8603828)%>% as_tibble%>% set_names("customer_id")
data_pick_four <- c(8618759, 8291696, 10357256, 8290374) %>% as_tibble %>% set_names("customer_id")
data_pick_five <- c(9044864, 8642053, 10534367, 9021526,11162275) %>% as_tibble %>% set_names("customer_id")
data_pick_six <- c(8221762, 8273636, 10359424, 8232822, 11450499)%>% as_tibble %>% set_names("customer_id")

@@

```{r assemble}
data_pick_cust <- bind_rows(
data_pick_one, 
#data_pick_two, 
data_pick_three,
data_pick_four,
data_pick_five, 
data_pick_six,
.id = "design") %>% 
 mutate(customer_id = as.character(customer_id))

@@

```{r data-pick}
data_pick <- read_rds(here::here("data/elec_nogap_2013_clean_356cust.rds")) %>%
  mutate(customer_id = as.character(customer_id)) %>% 
  dplyr::filter(customer_id %in% data_pick_cust$customer_id) %>% 
  gracsr::scale_gran( method = "robust",
                      response = "general_supply_kwh")
```

@@

```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in moy behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most households, there is one household for whom weekend behavior is standing out from the rest."}

data_hod <- quantile_gran(data_pick,
                                  "hour_day",
                                  quantile_prob_val = quantile_prob_graph) %>% 
  pivot_wider(names_from = quantiles,
              values_from = quantiles_values) %>% 
  left_join(data_pick_cust, by = c("customer_id"))

#data_heatmap_hod$customer_id = factor(data_heatmap_hod$customer_id, levels = data_pick_cust$value)
data_hod$category <- factor(data_hod$category, levels = 0:23)

hod_ind_design <- data_hod %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`,
                  group=customer_id),
              alpha = 0.5) +
  geom_line(aes(y = `50%`,
                group=customer_id), 
            size = 1) +
  facet_wrap(~customer_id, 
             scales = "free_y",
             nrow = 24) + 
 theme_application() +
  xlab("hod") +
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  scale_x_discrete(breaks = seq(0, 23, 3))

data_moy <- quantile_gran(data_pick,
                                  "month_year", 
                                  quantile_prob_val = quantile_prob_graph) %>% 
  pivot_wider(names_from = quantiles, 
              values_from = quantiles_values) %>% 
  left_join(data_pick_cust, by = c("customer_id"))

data_moy$category <- factor(data_moy$category, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

moy_ind_design <- data_moy %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`, 
                  group=customer_id), 
              alpha = 0.5) +
  geom_line(aes(y = `50%`, group=customer_id), size = 1) +
  facet_wrap(~customer_id, 
             scales = "free_y", 
             nrow = 24) +
    ylab("demand (in Kwh)") +
    xlab("moy")  +
  theme_application()

data_wkndwday <- data_pick  %>%
   create_gran("wknd_wday")  %>% 
  left_join(data_pick_cust, by = c("customer_id"))

ylim1 = boxplot.stats(data_wkndwday$general_supply_kwh)$stats[c(1, 5)]

wkndwday_ind_design <- data_wkndwday%>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(design), 
   #                   color = as.factor(design)), k=5, alpha = 0.5) +
  geom_boxplot(alpha = 0.5, fill = "black")+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
              nrow = 24)  +
  stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 1, color = "black")+
  xlab("wnwd")+
  theme_application()

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The distribution of electricity demand for the selected $24$ customers across `hod`, `moy` and `wnwd` are shown in Figure \ref{fig:hod-ind-group} (d, e, f). Our methodology is useful for grouping similar distributions over hod and moy and they are placed closely for easy comparison. Of course, certain customers in each group have distributions that differ from other members in the same group. However, it appears that the aim of grouping comparable distributions across categories of multiple temporal dependencies has been accomplished to some extent. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. It shows Groups $2$ and $4$ show a stronger hod pattern with a typical morning and evening peak, whereas groups 1, 3, and 5 show a moy pattern with higher usage in winter months. Differences in Weekend/Weekday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters unless maybe conditioned by `moy` or `hod`.

```{r all-data}
data_pick <- read_rds(here::here("data/elec_nogap_2013_clean_356cust.rds")) %>%
  mutate(customer_id = as.character(customer_id)) %>% 
  dplyr::filter(customer_id %in% data_pick_cust$customer_id) %>% 
  gracsr::scale_gran( method = "nqt",
                      response = "general_supply_kwh")
```

```{r clustering}
hod <- suppressMessages(data_pick %>% 
  dist_gran(gran1 = "hour_day", response = "general_supply_kwh"))

moy <- suppressMessages(data_pick %>% 
  dist_gran(gran1 = "month_year", response = "general_supply_kwh"))

wkndwday <- suppressMessages(data_pick %>% 
  dist_gran(gran1 = "wknd_wday", response = "general_supply_kwh"))

distance <- wkndwday/2 + moy/12 + hod/24

f = as.dist(distance)

@@

```{r opt-clusters}
library(fpc)
library(cluster)
k = array()
for(i in 2:20)
{
group <- f %>% hclust (method = "ward.D") %>% cutree(k=i)
p <- cluster.stats(f, clustering = group, silhouette = TRUE)
k[i]=p$sindex
}
# 
# ggplot(k %>% as_tibble %>% mutate(k = row_number()), aes(x=k, y = value)) + geom_line() + scale_x_continuous(breaks = seq(2, 20, 1))
```

```{r groups-24}
cluster_result <- suppressMessages(f %>% 
  clust_gran(kopt = 5)) %>% 
  rename("customer_id" = "id") %>% 
  mutate(group = as.factor(group))
```

```{r hod-ind-group, fig.cap="The distribution of selected consumers over hod (a, d), moy (b, e), and wkndwday (d, f). In each case, the same colour denotes the same group in plots (d), (e), (f) and are placed together to facilitate comparison. That means the customer orderings are different for (a, b, c) and (d, e, f). Our clustering methodology is useful for grouping similar distributions over hod and moy. Of course, certain customers in each group have distributions that differ from those of other members in the same group. However, it appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent." }

hod_ind_group <- data_hod %>% 
  left_join(cluster_result, by = c("customer_id")) %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`,
                  group=customer_id, fill = group),
              alpha = 0.5) +
  geom_line(aes(y = `50%`,
                group=customer_id, 
                color = group), size = 1) +
  facet_wrap(group~customer_id, 
             scales = "free_y",
              nrow=24) + 
  theme_application()+
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7")) +    xlab("hod")  +
  scale_x_discrete(breaks = seq(0, 23, 3))+ theme(legend.position = "bottom")
# + scale_y_continuous(breaks = NULL)

moy_ind_group <- data_moy %>% 
  left_join(cluster_result, by = c("customer_id")) %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`, 
                  group=customer_id, fill = group), alpha = 0.5) +
  geom_line(aes(y = `50%`, group=customer_id, color = group), size = 1) +
  facet_wrap(group~customer_id, 
             scales = "free_y", 
             nrow=24) +
    ylab("demand (in Kwh)") +
    xlab("moy")  +
  theme_application() +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7")) + theme(legend.position = "bottom")

wkndwday_ind_group <- data_wkndwday%>% 
  left_join(cluster_result, by = c("customer_id")) %>% 
  mutate(group  = as.factor(group)) %>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(group), 
          #            color = as.factor(group)), k=5, alpha = 0.5) +
  geom_boxplot(aes(color = group, fill = group),alpha = 0.5)+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(group~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
             nrow=24)  +
  theme_application()+
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7"))  + theme(legend.position = "none")+
  xlab("wnwd")

hod_ind_design + moy_ind_design + wkndwday_ind_design + hod_ind_group + moy_ind_group + wkndwday_ind_group +  plot_layout(guides = "collect", ncol = 6) + plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')') & theme(legend.position = 'none')
```

```{r data-heatmap-hod-group}
legend_title <- "group"

data_group <- data_pick  %>% 
  left_join(cluster_result, by = c("customer_id"))

data_heatmap_hod_group <- quantile_gran(data_group,
                                  gran1="hour_day",
                                  quantile_prob_val = c(0.25, 0.5, 0.75),
                                  group="group") %>% 
  pivot_wider(names_from = quantiles, values_from = quantiles_values)

data_heatmap_hod_group$category <- factor(data_heatmap_hod_group$category, levels = 0:23)

data_heatmap_hod_group$group <- paste("group", data_heatmap_hod_group$group, sep = "-")

hod_group <- data_heatmap_hod_group %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`,
                  group=group,
                  fill = as.factor(group), alpha = 0.5),
              alpha = 0.5) +
  geom_line(aes(y = `50%`,
                group=group, 
                color = as.factor(group)), size = 1)+
  facet_wrap(~group, 
             scales = "free_y",  
             nrow = 5) + 
              #labeller = labeller(xfacet = c(`1` = "Group 2", `2` = "Group 4",`3` = "Group 1",`4` = "Group 3"))
    theme(strip.text = element_text(size = 10, margin = margin(b = 0, t = 0))) + xlab("hour-of-day") + 
  ylab("demand (in Kwh)") + 
  theme_bw()  +
  scale_x_discrete(breaks = seq(1, 24, 3))+ 
  #theme(strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme_application() +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7")) +
  theme(legend.position = "bottom")

@@

```{r data-heatmap-moy-group}
data_heatmap_moy_group <- quantile_gran(data_group,
                                  gran1="month_year",
                                  quantile_prob_val = c(0.25, 0.5, 0.75),
                                  group="group") %>% 
  pivot_wider(names_from = quantiles, values_from = quantiles_values)

data_heatmap_moy_group$category <- factor(data_heatmap_moy_group$category, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

data_heatmap_moy_group$group <- paste("group", data_heatmap_moy_group$group, sep = "-")

moy_group <- data_heatmap_moy_group %>% 
  ggplot(aes(x = category)) + 
  geom_ribbon(aes(ymin = `25%`, 
                  ymax = `75%`, group=group, fill = as.factor(group)), alpha = 0.5) +
  geom_line(aes(y = `50%`, group=group, color = as.factor(group)), size = 1 ) +
  facet_wrap(~group, 
             scales = "free_y", 
             labeller = "label_value",
             nrow = 5) +
    theme(strip.text = element_text(size = 10, margin = margin(b = 0, t = 0))) + xlab("month-of-year") + 
  ylab("demand (in Kwh)") +
  theme_bw() + theme_application() +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7")) +
  theme(legend.position = "bottom") 
```

```{r}
wkndwday_data <- data_group %>% create_gran("wknd_wday") %>% 
  create_gran("hour_day")

ylim1 = boxplot.stats(wkndwday_data$general_supply_kwh)$stats[c(1, 5)]

wkndwday_group <- wkndwday_data%>% 
  ggplot(aes(x=hour_day, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(group)), k=5) +
  geom_boxplot(aes(fill = group, color = group),alpha = 0.5, outlier.alpha = 0.05)+
  #geom_boxplot(outlier.size = 1) + 
  coord_cartesian(ylim = ylim1*1.05)+
  #ggridges::geom_density_ridges2(aes(x = general_supply_kwh, y = wknd_wday,fill = as.factor(group))) + coord_flip() +
#geom_boxplot(aes(fill = as.factor(group))) +
  #scale_fill_lv() +
 xlab("wknd-wday") + 
  ylab("demand (in Kwh)") +
   facet_grid(group~wknd_wday, 
             scales = "free_y", 
             labeller = "label_both") + 
  theme_bw() + theme_application() +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7")) +
  theme(legend.position = "none") 
```

```{r combined-groups-js, fig.cap = "The distribution of electricity demand for the clusters across hod (a), moy (b) and wkndwday (c). It seems like group 2 and 5 have a hod pattern across its members, while group 1, 3, 5 have a moy pattern. Weekend/Weekday variations across groups are not distinguishable, indicating that it is not a critical variable for clustering. It is helpful to compare the summarised distributions of groups to that of individuals to confirm that the most of individuals in the group have the same characterisation."}

 (hod_group + moy_group + wkndwday_group) +
  plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')')+
  plot_layout(guides = "collect")& theme(legend.position = 'bottom')
```

@@

```{r data-pick-wpd}
elec_600_wpd <- read_rds(here::here("data/algo2-cust600-wpd-rawdata.rds"))

elec_pick <- elec_600_wpd %>% 
  filter(customer_id %in% data_pick_cust$customer_id)

elec_pick_wide <- elec_pick %>% pivot_wider(-c(1, 2), names_from = "x_variable", values_from = wpd)

scaled_var <- elec_pick_wide

f <- elec_pick_wide[-1] %>% dist()

k = array()
for(i in 2:20)
{
group <- f %>% hclust (method = "ward.D") %>% cutree(k=i)
p <- cluster.stats(f, clustering = group, silhouette = TRUE)
k[i]=p$sindex
}
# ggplot(k %>% as_tibble %>% mutate(k = row_number()), aes(x=k, y = value)) + geom_line() + scale_x_continuous(breaks = seq(2, 20, 1))

group <- f%>% hclust (method = "ward.D") %>% cutree(k=3)

cluster_result_wpd <- bind_cols(id = elec_pick_wide$customer_id, group = group)

data_pcp <- scaled_var %>% 
  #bind_cols(customer_id =  elec_pick_wide$customer_id) %>%
  left_join(cluster_result_wpd , by = c("customer_id" = "id")) %>% 
  select(customer_id, group, everything()) %>% 
  mutate(group = as.factor(group))

data_table <- data_pcp %>% group_by(group) %>% 
  summarise(nobs = n(),
            moy = round(median(month_year),2),
            hod = round(median(hour_day),2),
            wnwd = round(median(wknd_wday),2)) %>% 
  select(-group)

rownames(data_table) <- c("group-1", "group-2", "group-3")

@@

parcoord <- GGally::ggparcoord(data_pcp ,
                   columns = 3:ncol(data_pcp),
                   groupColumn = "group",
                   showPoints = FALSE, 
                   alphaLines = 0.8,
                   order = "anyClass",
                   scale = "globalminmax"
) + 
  ggplot2::theme(
    plot.title = ggplot2::element_text(size=10)
  )+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 10)) +
  theme(legend.position = "bottom") +
  xlab("") +
  ylab("wpd") + scale_fill_viridis_d(direction = 1) +
  scale_color_viridis_d(direction = 1) + theme_light()

(parcoord + gridExtra::tableGrob(data_table))+ plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')') & theme(legend.position = "bottom")

>>>>

```{r generate-design-3change}

@@

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature, yet it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection.
Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers who are far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that was hidden in a single variable display better than static projections. Hence we perform a linked tour with a t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to Supplementary article for further details). Figure \ref{fig:hod-ind-group} shows the distribution across `hod` (a), `moy`(b) and `wnwd` (c) for the set of chosen $24$ customers that were chosen. Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution.

```{r prototype-data-pick}

@@

```{r assemble}

@@

```{r data-pick}
```

@@

```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in moy behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most households, there is one household for whom weekend behavior is standing out from the rest."}
```

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The groupings are shown in Figure \ref{fig:hod-ind-group}. Customers with the same color represent the same clustered groups. Our methodology is useful for grouping similar distributions over `hod` and `moy` and they are placed closely for easy comparison. Few groups have mixed patterns across `hod` and `moy`, but few have all customers in the group having a similar profile. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. It shows Groups $2$ and $4$ have `hod` pattern with a typical morning and evening peak, whereas groups $1$, $3$, and $5$ show a `moy` pattern with higher usage in winter months. Differences in Weekend/Weekday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters unless maybe conditioned by `moy` or `hod`. It may be interesting to compare these two plots to verify if the summarized distributions across groups correctly characterised the groupings. If it has, then the majority of the group's members should share a similar profile.

<!--  It appears that the aim of grouping comparable distributions across categories of multiple temporal dependencies has been accomplished to some extent. -->

```{r all-data}
```

```{r clustering}

@@

```{r opt-clusters}
```

```{r groups-24}
```

```{r hod-ind-group, fig.cap = "The distribution of selected customers over hod (a), moy (b), and wnwd (c) for the $24$ selected customers is shown. In each case, the same colour denotes the same group using js-based clustering and are placed together to facilitate comparison. Individuals with similar distributional differences across categories of `hod` or `moy` are grouped together, with some exceptions. " }
```

<!-- It appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent. -->

```{r data-heatmap-hod-group}
```

```{r data-heatmap-moy-group}

@@

```{r data-heatmap-wkndwday-group}

```{r combined-groups-js, fig.cap = "The distribution of electricity demand over hod (a), moy (b), and wnwd*hod (c). Groups 2 and 5 appear to have a hod pattern among its members, whereas groups 1, 3, and 5 appear to have a moy pattern. Weekend/weekday differences across groups are not discernible, showing that it is not a critical variable for clustering on its own. It is probable that the distribution over Weekend/Weekday will differ conditional on the categories of hod or moy. (c) demonstrates that the distribution is not different, except for the tails. It is useful to compare the summarised distributions of groups to those of individuals in order to establish that the majority of individuals in the group share the same characteristics."}
```

@@

```{r data-pick-wpd}
```

@@
